\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-08-28 writing/manuscript.Rmd
started reviewing manuscript

---
title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r abstract, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change. Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century. We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses be interpreted holistically when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research, with particular attention to the study design, statistical modeling, and forecasting. We provide detailed `R` code and appendices so that other researchers can implement our methods on their own datasets."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

@@

<!--

other refs

* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3

-->

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016]. Additionally, temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and novel phenology [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust. We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.

@@

We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.

@@

Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource. A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average). Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures. Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods). Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- Appendix B$;$ @alston_mitigating_2022], while quadrature points had a weight of 1. Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. B13).

@@

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the area delimitated by the 99.9% Autocorrelated Densitity Kernel Estimates of space use. Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate SSPs in 2025. Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line. The projections only account for changes in movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four SSPs, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density."}
if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
knitr::include_graphics('../figures/local-distance-2100.png')
```

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS. Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that values < 1 indicate a decrease and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86)."}
if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
knitr::include_graphics('../figures/local-rss-2100.png')
```

@@

Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

\noindent The heterogeneity mammals' responses to temperature suggests that ecological communities will respond to change in complex and interconnected ways. Although our models do not account for explicit physiological or phenological changes, the surfaces in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears). For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day. Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer. The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020]. More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].

@@

The diversity in predicted changes in movement behavior by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. For example, the width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

>>>>

---
title: "\\Large The heat is on: Rising temperatures alter how and where mammals move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behavior cycles. Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century. We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less. Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes. Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effectively and proactively understanding mammalian movement behavior and protecting high-quality habitat. We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\

@@

# other refs

* @gerlich_species_2025: effects of temperature on fly movement in the arctic
* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].

Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018]. Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019]. Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust. We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.

@@

We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 reanalized data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.

@@

Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource. A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection [i.e., individual-level deviations from the species-level average$;$ @jeltsch_need_2025]. Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures. Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods). Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- Appendix B$;$ @alston_mitigating_2022], while quadrature points had a weight of 1. Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. B13).

@@

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within each species' observed extent. Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate SSPs in 2025. Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line. The projections only account for changes in movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios (i.e., not relative to other pixels). Values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density. All maps are shown with a BC Albers projection (EPSG:3005)."}
if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
knitr::include_graphics('../figures/local-distance-2100.png')
```

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in median relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within each species' observed extent within BC, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS. Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios (i.e., not relative to other pixels). Values < 1 indicate a decrease and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86). All maps are shown with a BC Albers projection (EPSG:3005)."}
if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
knitr::include_graphics('../figures/local-rss-2100.png')
```

@@

*Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.*

@@

\noindent The heterogeneity mammals' responses to temperature suggests that ecological communities will respond to change in complex and interconnected ways. Although our models do not account for explicit physiological or phenological changes, the surfaces in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears). For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day. Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer. The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across western Canada will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020]. More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].

@@

The diversity in predicted changes in movement behavior by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. For example, the width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025]. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
